---
title: "Are the Committees Everyone Calls 'Captured' Actually Preference Outliers?"
subtitle: "A pre-registered replication of Krehbiel (1990) on modern data, 103rd–117th Congress"
author: "Frank Frisbie"
date: today
format:
  html:
    output-file: report_v1.html
    theme: cosmo
    toc: true
    toc-location: left
    code-fold: true
    code-tools: true
    df-print: kable
    fig-width: 9
    fig-height: 4.5
execute:
  warning: false
  message: false
---

```{r setup}
library(dplyr); library(readr); library(ggplot2)
source("R/00_shared.R")

ideology <- read_csv("data/ideology.csv", show_col_types = FALSE)
assign   <- apply_icpsr_crosswalk(read_csv("data/committee_assignments_raw.csv",
                                           show_col_types = FALSE))
comp     <- completeness_table(assign, ideology)

# Per committee-Congress test results are produced (seeded) by R/03_analysis.R.
results   <- read_csv("data/committee_results.csv", show_col_types = FALSE)
agg_a     <- aggregate_test(results)
results_b <- results |> filter(!(chamber == "Senate" & congress %in% c(103, 108)))
agg_b     <- aggregate_test(results_b)

per_comm <- results |>
  group_by(group, chamber, label) |>
  summarise(outlier_rate = mean(outlier),
            mean_abs_disp = mean(abs_displacement),
            mean_signed_disp = mean(displacement), .groups = "drop")

pred_absrange <- per_comm |> filter(group == "predicted") |> pull(mean_abs_disp) |> range()
overall_rate  <- mean(results$outlier)
rules_disp    <- per_comm |> filter(label == "House Rules") |> pull(mean_abs_disp)
```

::: {.callout-note appearance="simple"}
**Archived first-pass report.** This is the standalone v1 write-up, kept for history. The canonical, up-to-date analysis — v1 *and* the sharper party-stratified v2 test, merged into one arc — is the [main page](index.html).
:::

## The claim being tested

One of the oldest ideas in the study of Congress is that committees are **captured**: that Agriculture is run by farm-state members, Armed Services by hawks and defense districts, the interior/lands committees by Western resource interests — each a cluster of **preference outliers** who don't look like the chamber as a whole. That's the distributive, "gains-from-trade" story (Shepsle–Weingast). Keith Krehbiel's *Are Committees Composed of Preference Outliers?* (APSR 1990) put it to a direct test and largely found **no**: committees look like the floor.

This is a small, deliberately narrow, **pre-registered** replication of that specific measurable claim on modern data (1993–2022): **do the committees distributive theory names sit measurably off the chamber median more often than chance, and more often than a control set?**

::: {.callout-note}
## This analysis was pre-registered
The hypothesis, the exact committee classification, the sample window, and the test were frozen in [`PRE_REGISTRATION.md`](PRE_REGISTRATION.md) **before any committee–ideology join was run or seen**. A result consistent with the null (committees mirror the chamber) was committed in advance to count as a finding, not a failure.
:::

## The headline result

**H1 is not supported.** The committees distributive theory says should be captured — Agriculture, Armed Services, Natural Resources / Energy, Transportation / Commerce — sit essentially **on top of** their chamber's median. Their committee medians are displaced from the floor by only about **`r sprintf("%.2f", pred_absrange[1])`–`r sprintf("%.2f", pred_absrange[2])`** on the −1…+1 DW-NOMINATE scale — tiny. Across every committee and Congress in scope, the permutation test flags a significant median outlier only **`r sprintf("%.1f%%", 100*overall_rate)`** of the time, *below* the 5% rate you'd expect by chance alone.

This is Krehbiel's 1990 result, replicated on 2020s data: **the committees everyone assumes are captured look like the chamber.**

```{r figure}
make_displacement_plot(results)
```

## Where the numbers come from

Every number traces to a public source, joined on ICPSR ID × Congress:

- **Committee rosters** — Stewart & Woon, *Congressional Committee Assignments* (Harvard Dataverse `congdata`): House `10.7910/DVN/XLIHUC`, Senate `10.7910/DVN/EQ6KC7`. Standing committees only; non-voting delegates excluded.
- **Member ideology** — voteview.com `HSall_members.csv`, DW-NOMINATE 1st dimension.

### Completeness was gated before any analysis

Following the discipline of the prior project, the pipeline computes the DW-NOMINATE match rate for committee members **per chamber-Congress and halts if any cell falls below 98%** — it refuses to analyze a partial join.

```{r completeness}
comp |>
  transmute(Chamber = chamber, Congress = congress, Members = members,
            Matched = matched, `Match rate` = round(match_rate, 4))
```

On the first run the gate did its job and **halted**: two Senate cells (103rd, 108th) came in at 97.9%. The cause was not missing data but a **linkage convention** — voteview assigns a member who switches party a separate 90000-series ICPSR for their post-switch ideal point, which the Stewart–Woon file does not mirror. Three members were affected, all with valid scores under the other ID:

```{r crosswalk}
ICPSR_CROSSWALK |>
  transmute(Member = member, State = state, Congress = congress,
            `Stewart-Woon ICPSR` = from_icpsr, `voteview ICPSR` = to_icpsr,
            `Party switch` = switch)
```

Mapping each to their verified voteview record is a data-linkage correction, not a researcher choice over the result (documented in the [`PRE_REGISTRATION.md`](PRE_REGISTRATION.md) deviation log). Lloyd Bentsen (103rd) is *not* crosswalked — he resigned days into the Congress and is genuinely scoreless, so the gate correctly excludes him. With the crosswalk applied, all 30 chamber-Congress cells clear 98%. Crucially, the aggregate result below is **identical whether the crosswalk is applied or the two affected Senate-Congress cells are dropped entirely** — so three people cannot be driving anything.

## The test, exactly as pre-registered

For each committee × Congress: compute the committee median and the chamber median (DW-NOMINATE dim 1); run a **two-sided, 10,000-draw permutation test** — draw random chamber subsets equal in size to the committee, and flag the committee as a *median outlier* if its observed median falls outside the central 95% of the null distribution of random-subset medians; record signed displacement (committee − chamber).

### Per-committee outlier rates

```{r per-committee}
per_comm |>
  transmute(Group = group, Committee = label,
            `Outlier rate` = round(outlier_rate, 3),
            `Mean signed disp.` = round(mean_signed_disp, 3),
            `Mean |disp.|` = round(mean_abs_disp, 3)) |>
  arrange(Group, desc(`Mean |disp.|`))
```

### Aggregate test of H1: predicted vs. not-predicted, reported both ways

```{r aggregate}
mk <- function(a, tag) tibble(
  Run = tag,
  `Outlier rate (pred)` = round(a$rate_pred, 3),
  `Outlier rate (not)`  = round(a$rate_not, 3),
  `Δ rate` = round(a$d_rate, 3),
  `Mean |disp| (pred)`  = round(a$md_pred, 3),
  `Mean |disp| (not)`   = round(a$md_not, 3),
  `Δ |disp|` = round(a$d_md, 3),
  `p (|disp|, 2-sided)` = round(a$p_md_twosided, 4))
bind_rows(
  mk(agg_a, "Crosswalk applied (full 103–117)"),
  mk(agg_b, "Senate 103 & 108 dropped")
)
```

The predicted-outlier set is a median outlier no more often than the control set (both ≈ `r sprintf("%.1f%%", 100*agg_a$rate_pred)`; the difference is `r sprintf("%+.3f", agg_a$d_rate)`, one-sided *p* = `r sprintf("%.2f", agg_a$p_rate_onesided)` — nowhere near supporting H1).

## An honest limitation: the null is permissive

The one aggregate difference that *is* significant runs **opposite** to H1: the predicted (constituency) committees are, if anything, slightly **more central** than the control set (mean |displacement| `r sprintf("%.3f", agg_a$md_pred)` vs. `r sprintf("%.3f", agg_a$md_not)`; two-sided *p* = `r sprintf("%.3f", agg_a$p_md_twosided)`). This is **not** evidence against distributive theory — it is an artifact of two things that must be named plainly:

1. **The permutation null is conservative by construction.** It draws *unconstrained* random subsets of the chamber. Real committees are built to roughly mirror the chamber's **party ratio**; unconstrained random subsets can be far more party-lopsided than any real committee, which inflates the spread of the null distribution. A wider null makes real committee medians look central and pushes the outlier rate *below* the 5% chance floor. The pre-registered **party-stratified null** (sampling within party delegations) is the sharper instrument and is the planned **v2**; the "not outliers" conclusion here should be read as holding *against a null deliberately tilted toward finding nothing*.

2. **The significant gap is driven by the control set, not the treatment set.** The largest displacement on the board belongs to **House Rules** (mean |displacement| `r sprintf("%.3f", rules_disp)`), which sits in the *not-predicted* group. Rules is unrepresentative because **party leadership packs it** — a party-structural effect, not a constituency-demand one. Rules was not part of the pre-registered focus, so this is an **exploratory observation, not a confirmatory finding**; it happens to widen the control set's average and thereby produce the opposite-signed significant difference.

## What this does and doesn't show

- **One statistic, one dimension.** This tests the *median* on DW-NOMINATE dim 1 — the classic quantity. Committees can differ from the floor in variance or on issue-specific intensity without a median shift; that is out of scope for v1.
- **The null matters.** As above, the unconstrained-random null is conservative; the party-stratified secondary (v2) is the sharper test and is pre-registered.
- **Exploratory vs. confirmatory.** The Rules/leadership observation is exploratory. The confirmatory result is the pre-registered predicted-vs-not comparison, which does not support H1.
- **Coverage ends 2022 (117th)** — the Stewart–Woon modern boundary.

## The one-line read

The committees everyone assumes are captured by narrow interests look like the floor; the one committee that looks unrepresentative here is the one leadership controls — House Rules — an exploratory observation, not a tested claim. Krehbiel-consistent, with a modern twist — and a sharper party-stratified test still to come.

## Reproduce this

```bash
Rscript R/01_pull_ideology.R        # DW-NOMINATE, 103-117
Rscript R/02_pull_committees.R      # Stewart-Woon rosters + delegate flags
Rscript R/03_analysis.R             # gate, permutation test, aggregate
quarto render report.qmd            # rebuild this page
```

Everything above regenerates from public data. Full frozen method and deviation log: [`PRE_REGISTRATION.md`](PRE_REGISTRATION.md).
